Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-09 mr-review.Rmd
added several refs

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
SNP is a genetic marker defined by nucleotide base identity at a single genomic position.

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable. The LD clumping procedure then selects high ranking SNPs that are not in high LD with each other.

LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by using a noncausal SNP that is in high LD with a 
causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
causal SNPs may be in high LD together. LD clumping is likely to omit one or more
of the causal SNPs.

@verbanck_detection_2018

Recent developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. GWAS summary statistics 
typically consist of an identifier and a p-value for every SNP in the data set, along with the sample 
size [@buniello2019nhgri].

@@

Investigators have found that combining information from multiple related 
traits can lead to richer MR results.

@@

Binary, survival, count - see discussion section of MRAID paper.

>>>>

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
tens of thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable. The LD clumping procedure then outputs high ranking SNPs that are not in high LD with each other.

LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by using a noncausal SNP that is in high LD with a 
causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
causal SNPs may be in high LD together. LD clumping is likely to omit one or more
of the causal SNPs, which can diminish statistical power in MR analysis.

Recent developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published has led to a rapid increase in the number of available
two-sample MR methods for GWAS summary statistics.

@@

Investigators have found that combining information from multiple related 
traits can lead to richer MR results [@liu_multi-trait_2021].

@@

Binary, survival, count - see discussion section of MRAID paper.
@allman_novel_2021 @burgess_improving_2012
